Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2024-12-29 LICS/sections/e-hypergraphs.tex
small edits

We do not expect internal interfaces to be preserved during rewriting: for example,  when the semilattice equations are considered as rewrites. 
Thus,  we wish a rewrite rule to be a pair of \textit{extended} cospans of e-hypergraphs with matching \textit{external} (but not necessarily internal) interfaces,  as follows. 
\[
    n \xrightarrow{} n' \xrightarrow{} \mathcal{L} \xleftarrow{} m' \xleftarrow{} m
\qquad
    n \xrightarrow{} n'' \xrightarrow{} \mathcal{R} \xleftarrow{} m'' \xleftarrow{} m
\] 
Analogous to Section \ref{sec:combinatorial-semantics}, observing that the following extended cospans express the same data
\[
    0 \xrightarrow{} 0 \xrightarrow{} \mathcal{L} \xleftarrow{} n'' + m'' \xleftarrow{} n + m
\qquad
    0 \xrightarrow{} 0 \xrightarrow{} \mathcal{R} \xleftarrow{} n' + m' \xleftarrow{} n + m
\] allows us to encode rewrite rules as extended cospans of the following form
\[
\mathcal{L} \xleftarrow{} n' + m' \xleftarrow{} n + m \xrightarrow{} n'' + m'' \xrightarrow{} \mathcal{R}
\] which will fit into the DPO formalism. We make this identification throughout,  without further mention. 
%We will futher omit the arrows $0 \to 0 \to \mathcal{L}$.
\begin{remark}
There is a limitation of extended DPOI rewriting as defined in this section: it is unable to express rewriting
for rules whose source or target contain a maximal connected sub-hypergraph with no input and output vertices.
While this limitation can be overcome by specifically encoding such rewrite rules, we find the technical development to be simpler if we avoid consideration of such rules and therefore consider signatures which can not result in construction of terms with type $0 \to 0$.
The discussion of a general case can be found in Section~\ref{sec:dpo-fix} of Appendix.
%This is because there is no way to impose both consistency and child relation of such an edge as it has no incident nodes in its context.
%We claim that this is only a minor limitation because in practice such edges (morphisms) with no
%incident nodes correspond to ‘scalars’, if we consider for example a category of vector spaces, and scalars are usually removed by being absorbed by vectors rather than introduced during rewrites.
\end{remark}

>>>>

We do not expect internal interfaces to be preserved during rewriting: for example,  when the semilattice equations are considered as rewrites. 
Thus,  we wish a rewrite rule to be a pair of \textit{extended} cospans of e-hypergraphs with matching \textit{external} (but not necessarily internal) interfaces,  as follows. 
\[
    n \xrightarrow{} n' \xrightarrow{} \mathcal{L} \xleftarrow{} m' \xleftarrow{} m
\qquad
    n \xrightarrow{} n'' \xrightarrow{} \mathcal{R} \xleftarrow{} m'' \xleftarrow{} m
\] 
Analogous to Section \ref{sec:combinatorial-semantics}, observing that the following extended cospans express the same data
\[
    0 \xrightarrow{} 0 \xrightarrow{} \mathcal{L} \xleftarrow{} n'' + m'' \xleftarrow{} n + m
\qquad
    0 \xrightarrow{} 0 \xrightarrow{} \mathcal{R} \xleftarrow{} n' + m' \xleftarrow{} n + m
\] allows us to encode rewrite rules as extended cospans of the following form
\[
\mathcal{L} \xleftarrow{} n' + m' \xleftarrow{} n + m \xrightarrow{} n'' + m'' \xrightarrow{} \mathcal{R}
\] which will fit into the DPO formalism. We make this identification throughout,  without further mention. 
%We will futher omit the arrows $0 \to 0 \to \mathcal{L}$.
\begin{remark}
The extended DPOI rewriting as defined in this section trades off simplicity for expressiveness: it is unable to express rewriting for rules whose source or target contain a maximal connected sub-hypergraph with no input and output vertices.
While this limitation can be overcome by specifically encoding such rewrite rules, we find the technical development to be simpler if we avoid consideration of such rules and therefore consider signatures which can not result in construction of terms with type $0 \to 0$. 
This limitation has no practical impact in applications of interest. 
However, the general case without this limitation can be found in Section~\ref{sec:dpo-fix} of Appendix.
%This is because there is no way to impose both consistency and child relation of such an edge as it has no incident nodes in its context.
%We claim that this is only a minor limitation because in practice such edges (morphisms) with no
%incident nodes correspond to ‘scalars’, if we consider for example a category of vector spaces, and scalars are usually removed by being absorbed by vectors rather than introduced during rewrites.
\end{remark}

### 2024-12-29 LICS/sections/introduction.tex
title
small edits

The other steps, (c) to (e), represent further elaborations of the e-graph, or e-hypergraph, via the application of more rules.
If we were to apply another rule, $x * 1 \to x$, we would identify the e-class of $a$ with the top-level e-class of the e-graph which would require a cycle to be introduced: the left argument of node $*$ would be the same e-class that $*$ belongs to.
This would result into a possibility of extracting the optimal term consisting of only the node $a$, discarding everything else.
To support such cycles we need a richer structure for our e-string diagrams.
While equipping the e-string diagrams which such structure is straightforward, it is out of the scope of this work, and thus we limit ourselves to acyclic e-graphs and e-string diagrams for the purposes of presentation.
Note that while the transformations of the e-graph are guided by the algorithm, the transformations of the e-string diagrammatic version is guided by the equations of the underlying theory.
A step-by-step application of such equations for the rewrite rule from $(a)$ to $(b)$ and $(b)$ to $(c)$ can be seen in Figure~\ref{fig:e-graph-example-a-b} and Figure~\ref{fig:e-graph-example-b-c} respectively.
The equations applied are the ones from Figure~\ref{fig:string-equations} and naturality of delete and copy.
One can note that the size of the eventual e-string diagram is comparable to the size of the corresponding e-graph.
% Note that in the final e-graph, (e), a cycle is introduced, making it the representation of infinitely many terms: $a, a*1, (a*1)*1, \ldots$. 
% From the saturated graph the optimal term consisting of only the node $a$ can be then extracted, discarding everything else.

@@

We will use enrichment to axiomatize the e-box structure as the ``join" of two morphisms $f,g: A \to B$ as $f + g: A \to B$. Note, in particular,  that semilattices do not require a unit for $+$. This is important, since, in any commutative monoid enriched category,  if there exists a categorical product, then it must be a categorical biproduct \cite{maclane}---a degeneracy we wish to avoid.  More precisely our hom-sets will be given the structure of a semilattice.  
Generally,  we may define a category \textit{enriched} in any other monoidal category $M$ (\textit{e.g.},  \textbf{SLatt}).  In this case,  hom-sets are generalised to hom-objects of $M$ (\textit{e.g.},  a semilattice) and composition 
\[
	\circ: Hom(B,C) \otimes Hom(A,B) \to Hom(A,C)
\]
is defined as an $M$-morphism.  In other words,  composition must respect the additional structure on hom-sets.  
In our case,  the category we wish to enrich is also monoidal,  so we additionally ask that the monoidal structure also respects this structure.  Technically, this amounts to asking that the tensor product is an enriched functor. 
We omit further details of the general case of enrichment here, and instead work with the following more concrete definition, where $M = \textbf{SLatt}$.

@@

Figure \ref{fig:egraph-strings} displays the generators of this language and Figure \ref{fig:string-equations} displays the additional equations which these diagrams satisfy, in addition to the standard SMC equations. 
The first four equations are those displayed in Definition \ref{def:enriched-prop},  while the final four axiomatize $+$ as an \textit{n-ary} associative, commutative and idempotent operation.  We overload the binary notation $+$ for our $n$-ary notation.  
It is intuitively clear that these diagrams are sound and complete with respect to their intended categorical semantics. 
Indeed, similar diagrammatic languages using boxes to express choice have been used before~\cite{duncan_generalised_2009}.

>>>>

The other steps, (c) to (e), represent further elaborations of the e-graph, or e-hypergraph, via the application of more rules.
If we were to apply another rule, $x * 1 \to x$, we would identify the e-class of $a$ with the top-level e-class of the e-graph which would require a cycle to be introduced: the left argument of node $*$ would be the same e-class that $*$ belongs to.
This would result into a possibility of extracting the optimal term consisting of only the node $a$, discarding everything else.
To support such cycles we need a richer structure for our e-string diagrams.
While equipping the e-string diagrams which such structure is straightforward, it is out of the scope of this work, and thus we limit ourselves to acyclic e-graphs and e-string diagrams for the purposes of presentation.
Note that while the transformations of the e-graph are guided by the algorithm, the transformations of the e-string diagrammatic version is guided by the equations of the underlying theory.
A detailed step-by-step application of the equations for the rewrite rules from $(a)$ to $(b)$ and $(b)$ to $(c)$ is given in Figures~\ref{fig:e-graph-example-a-b} and~\ref{fig:e-graph-example-b-c}, respectively, in the Appendix. 
The equations applied are the ones from Figure~\ref{fig:string-equations} along with the naturality of delete and copy.
One can note that the size of the eventual e-string diagram is comparable to the size of the corresponding e-graph.
% Note that in the final e-graph, (e), a cycle is introduced, making it the representation of infinitely many terms: $a, a*1, (a*1)*1, \ldots$. 
% From the saturated graph the optimal term consisting of only the node $a$ can be then extracted, discarding everything else.

@@

We will use enrichment to axiomatize the e-box structure as the ``join" of two morphisms $f,g: A \to B$ as $f + g: A \to B$. Note, in particular,  that semilattices do not require a unit for $+$. This is important, since, in any commutative monoid enriched category,  if there exists a categorical product, then it must be a categorical biproduct \cite{maclane}---a degeneracy we wish to avoid.  More precisely our hom-sets will be given the structure of a semilattice.  
Generally,  we may define a category \textit{enriched} in any other monoidal category $M$ (\textit{e.g.},  \textbf{SLatt}).  In this case,  hom-sets are generalised to hom-objects of $M$ (\textit{e.g.},  a semilattice) and composition 
\[
	\circ: Hom(B,C) \otimes Hom(A,B) \to Hom(A,C)
\]
is defined as an $M$-morphism.  In other words,  composition must respect the additional structure on hom-sets.  
In our case,  the category we wish to enrich is also monoidal,  so we additionally ask that the monoidal structure also respects this structure.  Technically, this amounts to asking that the tensor product is an enriched functor. 
We omit further details of the general case of enrichment here, and instead work with the concrete definition required for our application, where $M = \textbf{SLatt}$.
It remains to be seen if more general enrichments have interesting applications.

@@

Figure \ref{fig:egraph-strings} displays the generators of this language and Figure \ref{fig:string-equations} displays the additional equations which these diagrams satisfy, in addition to the standard SMC equations. 
The first four equations are those displayed in Definition \ref{def:enriched-prop},  while the final four axiomatize $+$ as an \textit{n-ary} associative, commutative and idempotent operation.  We overload the binary notation $+$ for our $n$-ary notation.  
We will later prove the intuitive fact that these diagrams are sound and complete with respect to their intended categorical semantics, noting that similar diagrammatic languages using boxes to express choice have been used before~\cite{duncan_generalised_2009}.
